---
title: "Supplementary code"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document provides all code for recreating the analyses published in Bird et al. Seeing the future. Where could invasive species eradication take us?

It builds on models developed for Bird J, Fuller RA, Terauds A, et al. 2021a. Detecting seabird responses to invasive species eradication. bioRxiv: 2021.04.07.438876.

We refer readers there for methods for model development and selection etc.

In places we adjust population estimates using occupancy data from Bird, J. P., Fuller, R. A., Pascoe, P. P., & Shaw, J. D. 2021. Using camera traps to determine occupancy and breeding in burrowing seabirds. bioRxiv. https://doi.org/10.1101/2021.04.08.438875

Load libraries for analyses.

```{r packages}
library(tidyverse) # for data minpulation and plotting
library(dsm) # for DSMs
library(sf) # for spatial analyses and plotting
library(raster) # for spatial analyses and plotting
library(dismo) # for BRTs
library(gbm) # for BRTs
library(scico) # for plot colours
library(ggnewscale) # for adjusting plot scales
#library(Distance)
#library(mgcv)
#library(mads)
#library(knitr)
#library(kableExtra)
#library(gridExtra)
#library(tabularaster)
```

Load data for analyses.

```{r}
load("data.RData")
```


## Presence-only models

### Antarctic Prion

Data come from Bird et al. (2021a). Observations from transects are filtered to remove all absences (zeros). Then a density surface model is built for the presence-only data using the dsm packacge - see Bird et al. (2021a) for full methods.

```{r presence-only ap}
# create presence-only data
seg_ap1 <- seg_ap %>% filter(Sample.Label %in% ap$Sample.Label)

# run models with different families to explore fit
# quasipoisson
dsm_ap_qp_pres <- dsm(count~s (dem, k=10) +
                    s (ndvi, k=10) +
                    s (ridge,k=10) +
                    s (slope, k=10) +
                    s (wetness, k=10) +
                    s (wind, k=10),
               ddf.obj=ap_detfc, segment.data=seg_ap1, 
               observation.data=ap, family=quasipoisson(), 
               segment.area=seg_ap1$area, keepData=TRUE, method="REML")
# par(mfrow=c(2, 2))
# gam.check(dsm_ap_qp_pres)
# summary(dsm_ap_qp_pres)

# negative-binomial
dsm_ap_nb_pres <- dsm(count~s (dem, k=10) +
                    s (ndvi, k=10) +
                    s (ridge,k=10) +
                    s (slope, k=10) +
                    s (wetness, k=10) +
                    s (wind, k=10),
               ddf.obj=ap_detfc, segment.data=seg_ap1, 
               observation.data=ap, family=nb(), 
               segment.area=seg_ap1$area, keepData=TRUE, method="REML")
# par(mfrow=c(2, 2))
# gam.check(dsm_ap_nb_pres)
# summary(dsm_ap_nb_pres)

# tweedie
dsm_ap_tw_pres <- dsm(count~s (dem, k=10) +
                    s (ndvi, k=10) +
                    s (ridge,k=10) +
                    s (slope, k=10) +
                    s (wetness, k=10) +
                    s (wind, k=10),
               ddf.obj=ap_detfc, segment.data=seg_ap1, 
               observation.data=ap, family=tw(), 
               segment.area=seg_ap1$area, keepData=TRUE, method="REML")
# par(mfrow=c(2, 2))
# gam.check(dsm_ap_tw_pres)
# summary(dsm_ap_tw_pres)

# negative-binomial model has lowest AIC and diagnostics suggest a marginally better fit
```

```{r model testing ap}
dsm_ap_nb_pres1 <- dsm(count~s (dem, k=10) +
                    s (ndvi, k=10) +
                    s (slope, k=10) +
                    s (wetness, k=10) +
                    s (wind, k=10),
               ddf.obj=ap_detfc, segment.data=seg_ap1, 
               observation.data=ap, family=nb(), 
               segment.area=seg_ap1$area, keepData=TRUE, method="REML") # drop ridge (highest non-signif p-values)
# par(mfrow=c(2, 2))
# gam.check(dsm_ap_nb_pres1)
# summary(dsm_ap_nb_pres1)

dsm_ap_nb_pres2 <- dsm(count~s (dem, k=10) +
                    s (ndvi, k=10) +
                    s (slope, k=10) +
                    s (wind, k=10),
               ddf.obj=ap_detfc, segment.data=seg_ap1, 
               observation.data=ap, family=nb(), 
               segment.area=seg_ap1$area, keepData=TRUE, method="REML") # drop wetness (highest non-signif p-value)
# par(mfrow=c(2, 2))
# gam.check(dsm_ap_nb_pres2)
# summary(dsm_ap_nb_pres2)

# all terms coming out as significant.
```

```{r summarise dsms function}
summarize_dsm <- function(model){

  summ <- summary(model)

  data.frame(response = model$family$family,
             terms    = paste(rownames(summ$s.table), collapse=", "),
             "Adjusted R.sq" = round(summ$r.sq,2),
             AIC      = AIC(model),
             REML     = model$gcv.ubre,
             "Deviance_explained" = paste0(round(summ$dev.expl*100,2),"%")
            )

}
```

```{r dsm selection ap}
dsms_ap <- mget(ls(pattern = "dsm_ap"))
tab_ap <- plyr::ldply(dsms_ap, summarize_dsm)
# simpler nb model has lowest AIC
```

Having selected a model now predict density across Macquarie Island.

```{r prediction ap}
mac$area <- 400

over_pred_ap <- predict(dsm_ap_nb_pres2, mac, off.set=mac$area)

# estimate the total number of burrows predicted island-wide
brw_oest_ap <- sum(over_pred_ap, na.rm = T)
over_pred_apd <- over_pred_ap/400 # turn the estimated abundance into density

# calculate uncertainty using the delta method
mac_var <- mac[!is.na(mac$ndvi),]
over_pred_ap_var <- dsm.var.gam(dsm_ap_nb_pres2, pred.data = mac_var, off.set=mac_var$area)
over_ap_ci <- summary(over_pred_ap_var) # simple uncertainty

# setup the storage for the predictions
over_ap_raster <- raster(sta)
# put the values in, making sure they are numeric first
over_ap_raster <- setValues(over_ap_raster, as.numeric(over_pred_apd))
# name the new, last, layer in the stack
names(over_ap_raster) <- "dsm_ap_nb_pres2"

```

And use the modelled density to calculate a projected population estimate corrected for occupancy (as measured during surveys in 2017-2018).

```{r ap presence-only model pop estimate}
# correct brw est for occupancy
pop_oest_ap <- sum(over_pred_ap*ap_occ, na.rm = T)

# work out uncertainty using delta method to combine burrows and occupancy:
# get CV of occ and brw ests
ap_occ_cv <- ap_occ_se/ap_occ
brw_oest_ap_cv <- summary(over_pred_ap_var)$cv
# combine cv's
pop_oest_ap_cv <- sqrt((brw_oest_ap_cv^2) + (ap_occ_cv^2))

# 95% CIs for pop est
var.D <- (pop_oest_ap_cv*pop_oest_ap)^2
var.logD <- log(1+(var.D/(pop_oest_ap^2)))

c.value <- exp(1.96*(sqrt(var.logD)))

# upper and lower CIs
pop_oest_ap_upr <- pop_oest_ap*c.value
pop_oest_ap_lwr <- pop_oest_ap/c.value
```


### White-headed Petrel

Data come from Bird et al. (2021a). Observations from transects are filtered to remove all absences (zeros). Then a density surface model is built for the presence-only data using the dsm packacge - see Bird et al. (2021a) for full methods.

```{r presence-only whp}
seg_whp1 <- seg_whp %>% filter(Sample.Label %in% whp$Sample.Label)

# run models with different families to explore fit
# quasipoisson
dsm_whp_qp_pres <- dsm(count~s (dem, k=10) +
                    s (ndvi, k=10) +
                    s (ridge,k=10) +
                    s (slope, k=10) +
                    s (wetness, k=10) +
                    s (wind, k=10),
               ddf.obj=whp_detfc, segment.data=seg_whp1, 
               observation.data=whp, family=quasipoisson(), 
               segment.area=seg_whp1$area, keepData=TRUE, method="REML")
# par(mfrow=c(2, 2))
# gam.check(dsm_whp_qp_pres)
# summary(dsm_whp_qp_pres)

# negative-binomial
dsm_whp_nb_pres <- dsm(count~s (dem, k=10) +
                    s (ndvi, k=10) +
                    s (ridge,k=10) +
                    s (slope, k=10) +
                    s (wetness, k=10) +
                    s (wind, k=10),
               ddf.obj=whp_detfc, segment.data=seg_whp1, 
               observation.data=whp, family=nb(), 
               segment.area=seg_whp1$area, keepData=TRUE, method="REML")
# par(mfrow=c(2, 2))
# gam.check(dsm_whp_nb_pres)
# summary(dsm_whp_nb_pres)

# tweedie
dsm_whp_tw_pres <- dsm(count~s (dem, k=10) +
                    s (ndvi, k=10) +
                    s (ridge,k=10) +
                    s (slope, k=10) +
                    s (wetness, k=10) +
                    s (wind, k=10),
               ddf.obj=whp_detfc, segment.data=seg_whp1, 
               observation.data=whp, family=tw(), 
               segment.area=seg_whp1$area, keepData=TRUE, method="REML")
# par(mfrow=c(2, 2))
# gam.check(dsm_whp_tw_pres)
# summary(dsm_whp_tw_pres)

# negative-binomial model gives the best fit to the data and lowest AIC
```

```{r model testing whp}
dsm_whp_nb_pres1 <- dsm(count~s (dem, k=10) +
                    s (ndvi, k=10) +
                    s (slope, k=10) +
                    s (wetness, k=10) +
                    s (wind, k=10),
               ddf.obj=whp_detfc, segment.data=seg_whp1, 
               observation.data=whp, family=nb(), 
               segment.area=seg_whp1$area, keepData=TRUE, method="REML") # drop ridge (highest non-signif p-values)
# par(mfrow=c(2, 2))
# gam.check(dsm_whp_nb_pres1)
# summary(dsm_whp_nb_pres1)

dsm_whp_nb_pres2 <- dsm(count~s (dem, k=10) +
                    s (slope, k=10) +
                    s (wetness, k=10) +
                    s (wind, k=10),
               ddf.obj=whp_detfc, segment.data=seg_whp1, 
               observation.data=whp, family=nb(), 
               segment.area=seg_whp1$area, keepData=TRUE, method="REML") # drop wetness (highest non-signif p-value)
# par(mfrow=c(2, 2))
# gam.check(dsm_whp_nb_pres2)
# summary(dsm_whp_nb_pres2)

dsm_whp_nb_pres3 <- dsm(count~s (dem, k=10) +
                    s (slope, k=10) +
                    s (wetness, k=10),
               ddf.obj=whp_detfc, segment.data=seg_whp1, 
               observation.data=whp, family=nb(), 
               segment.area=seg_whp1$area, keepData=TRUE, method="REML") # drop wetness (highest non-signif p-value)
# par(mfrow=c(2, 2))
# gam.check(dsm_whp_nb_pres3)
# summary(dsm_whp_nb_pres3)

# all terms coming out as significant.
```

```{r dsm selection whp}
dsms_whp <- mget(ls(pattern = "dsm_whp"))
tab_whp <- plyr::ldply(dsms_whp, summarize_dsm)
# simplest model has lowest AIC - use for predicting
```

Having selected a model now predict density across Macquarie Island.

```{r prediction whp}
over_pred_whp <- predict(dsm_whp_nb_pres3, mac, off.set=mac$area)

# estimate the total number of burrows predicted island-wide
brw_oest_whp <- sum(over_pred_whp, na.rm = T)
over_pred_whpd <- over_pred_whp/400 # turn the estimated abundance into density

# calculate uncertainty using the delta method
mac_var <- mac[!is.na(mac$ndvi),]
over_pred_whp_var <- dsm.var.gam(dsm_whp_nb_pres3, pred.data = mac_var, off.set=mac_var$area)
# pred_whp_varprop <- dsm.var.prop(dsm_whp_nb3, pred.data = mac_var, off.set=mac_var$area)
over_whp_ci <- summary(over_pred_whp_var) # simple uncertainty

# setup the storage for the predictions
over_whp_raster <- raster(sta)
# put the values in, making sure they are numeric first
over_whp_raster <- setValues(over_whp_raster, as.numeric(over_pred_whpd))
# name the new, last, layer in the stack
names(over_whp_raster) <- "dsm_whp_nb_pres3"

```

And use the modelled density to calculate a projected population estimate corrected for occupancy (as measured during surveys in 2017-2018).

```{r whp presence-only model pop estimate}
# correct brw est for occupancy
pop_oest_whp <- sum(over_pred_whp*whp_cocc, na.rm = T)

# work out uncertainty:
# get CV of occ and brw ests
whp_cocc_cv <- whp_cocc_se/whp_cocc
brw_oest_whp_cv <- summary(over_pred_whp_var)$cv
# combine cv's
pop_oest_whp_cv <- sqrt((brw_oest_whp_cv^2) + (whp_cocc_cv^2))

# 95% CIs for pop est
var.D <- (pop_oest_whp_cv*pop_oest_whp)^2
var.logD <- log(1+(var.D/(pop_oest_whp^2)))

c.value <- exp(1.96*(sqrt(var.logD)))

# upper and lower CIs
pop_oest_whp_upr <- pop_oest_whp*c.value
pop_oest_whp_lwr <- pop_oest_whp/c.value
```


### Blue Petrel

Data come from Bird et al. (2021a). Observations from a targeted whole-island survey were aggregated within cells of a whole-island grid, and filtered to remove all cells with no observations - absences (zeros). Then gams are built for the presence-only data using the mgcv packacge - see Bird et al. (2021a) for full methods.

```{r bp presence model}
dat_bpp <- dat_bp %>% filter(ct>0)

bpAp <- gam(ct ~  offset(log(area)) +
              s(dem) + 
              s(ndvi) +
              s(ridge) +
              s(slope) +
              s(wetness) +
              s(aspect) +
              s(wind), data = dat_bpp, family = nb())
# summary(bpAp)
# plot(bpAp, pages = 1)
# par(mfrow=c(2, 2))
# gam.check(bpAp)
```

Use the model to predict density across Macquarie Island.

```{r bp pres predictions}
#library(raster)
gd <- as.data.frame(sta, xy = TRUE)
gd$area <- 400 # figure out the offset
bpAp_p <- predict(bpAp, newdata = gd, type = "response", se.fit = TRUE)
bpp_est <- (sum(bpAp_p$fit, na.rm=T)) # count per pixel
bpp_min <- bpp_est - ((sum(bpAp_p$se.fit, na.rm=T))*1.96)
bpp_max <- bpp_est + ((sum(bpAp_p$se.fit, na.rm=T))*1.96)
bpAp_pd <- bpAp_p$fit/400 # converting into density of burrows per m^2

bpAp_pr <- raster(sta)
gd_bpp <- gd
gd_bpp$cells <- cellFromXY(bpAp_pr, gd_bpp[c("x", "y")])
bpAp_pr[gd_bpp$cells] <- bpAp_pd
# par(mfrow=c(1, 1))
# plot(bpAp_pr)

#detach("package:raster", unload = TRUE)
```

And use the modelled density to calculate a projected population estimate corrected for occupancy (as measured during surveys in 2017-2018).

```{r bp presence-only model pop estimate}
# correct brw est for occupancy
pop_oest_bp <- sum(bpAp_p$fit*bp_occ, na.rm = T)

# work out uncertainty:
# get CV of occ and brw ests
bp_occ_cv <- bp_occ_se/bp_occ
brw_oest_bp_cv <- (sum(bpAp_p$se.fit, na.rm = T)/bpp_est)
# combine cv's
pop_oest_bp_cv <- sqrt((brw_oest_bp_cv^2) + (bp_occ_cv^2))

# 95% CIs for pop est
var.D <- (pop_oest_bp_cv*pop_oest_bp)^2
var.logD <- log(1+(var.D/(pop_oest_bp^2)))

c.value <- exp(1.96*(sqrt(var.logD)))

# upper and lower CIs
pop_oest_bp_upr <- pop_oest_bp*c.value
pop_oest_bp_lwr <- pop_oest_bp/c.value
```


### Grey Petrel

Data come from Bird et al. (2021a). Observations from a targeted whole-island survey were aggregated within cells of a whole-island grid, and filtered to remove all cells with no observations - absences (zeros). Then gams are built for the presence-only data using the mgcv packacge - see Bird et al. (2021a) for full methods.

```{r gp presence model}
dat_gpp <- dat_gp %>% filter(ct>0)

gpAp <- gam(ct ~  offset(log(area)) +
              s(dem) + 
              s(ndvi) +
              s(ridge) +
              s(slope) +
              s(wetness) +
              s(aspect) +
              s(wind), data = dat_gpp, family = nb())
# summary(gpAp)
# plot(gpAp, pages = 1)
# par(mfrow=c(2, 2))
# gam.check(gpAp)
```

```{r}
# library(raster)
# gd <- as.data.frame(sta, xy = TRUE)
# gd$area <- 400 # figure out the offset
gpAp_p <- predict(gpAp, newdata = gd, type = "response", se.fit = TRUE)
gpp_est <- (sum(gpAp_p$fit, na.rm=T)) # count per pixel
gpp_min <- gpp_est - ((sum(gpAp_p$se.fit, na.rm=T))*1.96)
gpp_max <- gpp_est + ((sum(gpAp_p$se.fit, na.rm=T))*1.96)
gpAp_pd <- gpAp_p$fit/400 # converting into density of burrows per m^2

gpAp_pr <- raster(sta)
gd_gpp <- gd
gd_gpp$cells <- cellFromXY(gpAp_pr, gd_gpp[c("x", "y")])
gpAp_pr[gd_gpp$cells] <- gpAp_pd
# par(mfrow=c(1, 1))
# plot(gpAp_pr)
# 
# detach("package:raster", unload = TRUE)
```

```{r ap presence-only model pop estimate}
# correct brw est for occupancy
pop_oest_gp <- sum(gpAp_p$fit*gp_cocc, na.rm = T)

# work out uncertainty:
# get CV of occ and brw ests
gp_cocc_cv <- gp_cocc_se/gp_cocc
brw_oest_gp_cv <- (sum(gpAp_p$se.fit, na.rm = T)/gpp_est)
# combine cv's
pop_oest_gp_cv <- sqrt((brw_oest_gp_cv^2) + (gp_cocc_cv^2))

# 95% CIs for pop est
var.D <- (pop_oest_gp_cv*pop_oest_gp)^2
var.logD <- log(1+(var.D/(pop_oest_gp^2)))

c.value <- exp(1.96*(sqrt(var.logD)))

# upper and lower CIs
pop_oest_gp_upr <- pop_oest_gp*c.value
pop_oest_gp_lwr <- pop_oest_gp/c.value
```


### Figure 2

Figure 2 overlays the current distributions of all species (taken from Bird et al. 2021a) in the first panel, and then plots the presence-only modelled predictions made above in the remaining panels, one for each species.

```{r adjusting lat/long scales}
scale_x_longitude <- function(xmin=-180, xmax=180, step=1, ...) {
    ewbrks <- seq(xmin,xmax,step)
    ewlbls <- unlist(lapply(ewbrks, function(x) ifelse(x < 0, paste(x, "W"), ifelse(x > 0, paste(x, "E"),x))))
    return(scale_x_continuous("Longitude", breaks = ewbrks, labels = ewlbls, expand = c(0, 0), ...))
}
scale_y_latitude <- function(ymin=-90, ymax=90, step=0.5, ...) {
    nsbrks <- seq(ymin,ymax,step)
    nslbls <- unlist(lapply(nsbrks, function(x) ifelse(x < 0, paste(x, "S"), ifelse(x > 0, paste(x, "N"),x))))
    return(scale_y_continuous("Latitude", breaks = nsbrks, labels = nslbls, expand = c(0, 0), ...))
}
```

```{r coast}
# coast layer
crs <- st_crs(ccoast)
ccoast_buff <- st_buffer(ccoast, dist = 800)
```

Panel 1, current distribution of all species.

```{r panel 1}
# filter only pixels with >1 breeding pair predicted
pr_apf <- pr_ap %>% filter(pc > 1)
pr_whpf <- pr_whp %>% filter(pc > 1)

# normalise densities for all spp.
pr_apr <- pr_apf %>% filter(pd > quantile(pd, 0.025) & pd < quantile(pd, 0.975)) %>% 
         mutate(pd_norm = (pd-min(pd))/diff(range(pd)))
pr_whpr <- pr_whpf %>% filter(pd > quantile(pd, 0.025) & pd < quantile(pd, 0.975)) %>% 
           mutate(pd_norm = (pd-min(pd))/diff(range(pd)))
gdat_gpr <- gdat_gp %>% filter(dens > quantile(dens, 0.025) & dens < quantile(dens, 0.975)) %>% 
         mutate(pd_norm = (dens-min(dens))/diff(range(dens)))
gdat_bpr <- gdat_bp %>% filter(dens > quantile(dens, 0.025) & dens < quantile(dens, 0.975)) %>% 
         mutate(pd_norm = (dens-min(dens))/diff(range(dens)))

ggplot() +
    # set-up plot background etc.
    geom_sf(fill = 'grey25', data = ccoast) +
    geom_sf(fill = NA, colour = NA, data = ccoast_buff) +
    scale_x_longitude(158.8, 158.9, step = 0.1) +
    # add GP layer #440154FF #E0754FFF
    geom_sf(data = gdat_gpr, aes(col = pd_norm), show.legend = T) +
    scale_color_gradient(low = "transparent", high = "#440154FF", aesthetics = 'colour') +
    # add BP layer #33638DFF #2A5934FF
    new_scale_colour() +
    geom_sf(data = gdat_bpr, aes(col = pd_norm), show.legend = T) +
    scale_color_gradient(low = "transparent", high = "#33638DFF", aesthetics = 'colour') +
    # add WHP layer #3CBB75FF #2B5B99FF
    new_scale_fill() +
    geom_raster(aes(x=x, y=y, fill=pd_norm), data = pr_whpr) +
    scale_fill_gradient(low = "transparent", high = "#3CBB75FF", aesthetics = 'fill') +
    # add AP layer #FDE725FF #C36D9AFF
    new_scale_fill() +
    geom_raster(aes(x=x, y=y, fill=pd_norm), data = pr_apr) +
    scale_fill_gradient(low = "transparent", high = "#FDE725FF", aesthetics = 'fill') +
    theme_bw()
```

Projected distributions based upon presence-only models - individual species maps

```{r normalise presence-only predictions}
# normalise projected distribution layer for AP
pr_fap <- as.data.frame(over_ap_raster, xy = T) %>% 
         drop_na() %>% 
         filter(dsm_ap_nb_pres2 > quantile(dsm_ap_nb_pres2, 0.025) & 
                dsm_ap_nb_pres2 < quantile(dsm_ap_nb_pres2, 0.975)) %>% 
         mutate(pd_norm = (dsm_ap_nb_pres2-min(dsm_ap_nb_pres2))/diff(range(dsm_ap_nb_pres2))) 

# normalise projected distribution layer for AP
pr_fwhp <- as.data.frame(over_whp_raster, xy = T) %>% 
         drop_na() %>% 
         filter(dsm_whp_nb_pres3 > quantile(dsm_whp_nb_pres3, 0.025) & 
                dsm_whp_nb_pres3 < quantile(dsm_whp_nb_pres3, 0.975)) %>% 
         mutate(pd_norm = (dsm_whp_nb_pres3-min(dsm_whp_nb_pres3))/diff(range(dsm_whp_nb_pres3))) 

# normalise projected distribution layer for AP
pr_fgp <- as.data.frame(gpAp_pr, xy = T) %>% 
         drop_na() %>% 
         filter(layer > quantile(layer, 0.025) & 
                layer < quantile(layer, 0.975)) %>% 
         mutate(pd_norm = (layer-min(layer))/diff(range(layer))) 

# normalise projected distribution layer for BP
pr_fbp <- as.data.frame(bpAp_pr, xy = T) %>% 
         drop_na() %>% 
         filter(layer > quantile(layer, 0.025) & 
                layer < quantile(layer, 0.975)) %>% 
         mutate(pd_norm = (layer-min(layer))/diff(range(layer))) 
```

```{r species maps}
ggplot() +
    # set-up plot background etc.
    geom_sf(fill = 'grey25', data = ccoast) +
    geom_sf(fill = NA, colour = NA, data = ccoast_buff) +
    scale_x_longitude(158.8, 158.9, step = 0.1) +
    # add AP layer #FDE725FF #C36D9AFF
    geom_raster(aes(x=x, y=y, fill=pd_norm), data = pr_fap) +
    scale_fill_gradient(low = "transparent", high = "#FDE725FF", aesthetics = 'fill') +
    theme_bw()

ggplot() +
    # set-up plot background etc.
    geom_sf(fill = 'grey25', data = ccoast) +
    geom_sf(fill = NA, colour = NA, data = ccoast_buff) +
    scale_x_longitude(158.8, 158.9, step = 0.1) +
    # add GP layer #440154FF #E0754FFF
    new_scale_fill() +
    geom_raster(aes(x=x, y=y, fill=pd_norm), data = pr_fgp) +
    scale_fill_gradient(low = "transparent", high = "#440154FF", aesthetics = 'fill') +
    theme_bw()

ggplot() +
    # set-up plot background etc.
    geom_sf(fill = 'grey25', data = ccoast) +
    geom_sf(fill = NA, colour = NA, data = ccoast_buff) +
    scale_x_longitude(158.8, 158.9, step = 0.1) +
    # add WHP layer #3CBB75FF #2B5B99FF
    new_scale_fill() +
    geom_raster(aes(x=x, y=y, fill=pd_norm), data = pr_fwhp) +
    scale_fill_gradient(low = "transparent", high = "#3CBB75FF", aesthetics = 'fill') +
    theme_bw()

ggplot() +
    # set-up plot background etc.
    geom_sf(fill = 'grey25', data = ccoast) +
    geom_sf(fill = NA, colour = NA, data = ccoast_buff) +
    scale_x_longitude(158.8, 158.9, step = 0.1) +
    # add BP layer #33638DFF #2A5934FF
    new_scale_fill() +
    geom_raster(aes(x=x, y=y, fill=pd_norm), data = pr_fbp) +
    scale_fill_gradient(low = "transparent", high = "#33638DFF", aesthetics = 'fill') +
    theme_bw()
```


## Inter-island comparison

The second part of our analysis uses data on Blue Petrels from different islands to model their distribution on each island, and use this model to predict their distribution and abundance on Macquarie, if they occupied the same niche, and at the same density as on comparative islands.

```{r Data exploration1}
mean(all$dens)
max(all$dens)
nrow(all)
pois <- data.frame(6954*dpois(0:5, 0.04, log = FALSE))
names(pois)[1] <- "pois"
x <- 0:5
pois$levels <- factor(x)

dist <- all %>% group_by(group = cut(dens, breaks = seq(0, max(dens), 1))) %>%
  summarise(n = n()) %>% mutate(levels = factor(c(1,2,3,4,0)))
pois <- pois %>% left_join(dist, by = "levels")
pois <- dplyr::select(pois, levels, pois, n)
pois[is.na(pois)] <- 0
pois$pois <- as.integer(pois$pois)
pois <- pivot_longer(pois, c(pois, n), values_to = "count")

# data don't look zero-inflated, but are probably over-dispersed.
ggplot(pois, aes(levels, count)) +
  geom_bar(stat = "identity", position = "dodge", aes(fill = name))

```

```{r Data exploration2}
# Correlation panel
panel.cor <- function(x, y){
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    r <- round(cor(x, y), digits=2)
    txt <- paste0("R = ", r)
    cex.cor <- 0.8/strwidth(txt)
    text(0.5, 0.5, txt, cex = cex.cor * r)
}
# Customize upper panel
upper.panel<-function(x, y){
  points(x,y, pch = 19)
}
# Pairs plot to examine correlation between factors - BRTs require data that is not highly correlated.
pairs(all[,c(6:12)], 
      lower.panel = panel.cor,
      upper.panel = upper.panel)

# Compare distribution in relation to environmental variables between the three islands
ggplot(corr) +
         geom_histogram(aes(val, weight=dens)) +
         facet_grid(vars(island), vars(var), scales="free", shrink = T) +
         xlab("Covariate value") +
         ylab("Density")
```

```{r load island grids}
cgrid$isl1 <- factor("macca")
cgrid$isl2 <- factor("marion")
cgrid$isl3 <- factor("mayes")
```

### Boosted regression trees

Diagnostic plots for hierarchical GAMs fit to the pooled data were poor (Pedersen et al. 2019), so in addition we used Boosted Regression Trees (BRT) which behave like an additive regression model. Previous studies show BRTs have high stability and predictive power capable of dealing with patchy datasets with no need for data transformation or removal of outliers (Elith et al. 2008; Guillaumot et al. 2018). We fitted two BRTs and combined them using a delta-lognormal approach (Dedman et al. 2017). BRT1 modelled presence-absence, with predictions on a continuous scale between zero and one being converted into binary predictions maximizing the sum of sensitivity and speciﬁcity (max SSS) as a threshold (Liu et al. 2013). This binary layer was then used as a mask for spatial predictions of abundance from BRT2, a presence-only model of logged density (Dedman et al. 2017). BRTs were built with a Bernoulli distribution, tree complexity = 3, learning rate = 0.01 and bag fraction = 0.5. Models were run for >1,000 trees and evaluated using 10-fold cross validation, then checked for deviance and area under the receiver operating characteristic curve (Elith et al., 2008). ‘Island’ was included as a factor in the BRTs, and inter-island predictions were made by simply renaming the Macquarie Island spatial environmental prediction data as “Marion” or “Mayes”. 

```{r}
all <- all %>% mutate(pres = case_when(dens > 0 ~ 1,
                                       TRUE ~ 0),
                      log_dens = log(dens))
```

```{r step 1}
# step 1 - model presence-absence. Uses bernoulli (= binomial) model
# stratify by island
island <- all %>% group_by(island) %>% count()
n_folds <- 3 # number of islands
fold_by_island <- all %>% count(island) %>% arrange(desc(n)) %>% 
                  mutate(fold = rep(seq_len(n_folds),
                                    ceiling(n()/n_folds))[seq_len(n())])
all <- left_join(all, dplyr::select(fold_by_island, island, fold), by = "island")

# BRT model with island as a factor
brt1 <- gbm.step(data=all, gbm.x = c(1,6:11), gbm.y = 13,
                        family = "bernoulli", tree.complexity = 3,
                        learning.rate = 0.01, bag.fraction = 0.5)

# # BRT model with island for cross-validation
# brt2 <- gbm.step(data=all, gbm.x = c(6:11), gbm.y = 13,
#                 family = "bernoulli", tree.complexity = 3, 
#                 n.folds = n_folds, fold.vector = all$fold,
#                 learning.rate = 0.00005, bag.fraction = 0.5)

```

```{r}
# BRT density model with island as a factor
brt3 <- gbm.step(data=filter(all, dens > 0), gbm.x = c(1,6:11), gbm.y = 14,
                        family = "gaussian", tree.complexity = 3,
                        learning.rate = 0.01, bag.fraction = 0.5)

# BRT density model with island for cross-validation
# brt4 <- gbm.step(data=filter(all, dens > 0), gbm.x = c(6:11), gbm.y = 14,
#                 family = "gaussian", tree.complexity = 3, 
#                 n.folds = n_folds, fold.vector = all$fold,
#                 learning.rate = 0.001, bag.fraction = 0.5)

```

#### Inter- and intra-island predictions
##### Presence-absence

First use the presence-absence model to predict presence-absence island-wide

```{r macca to macca}
cgrid$island <- cgrid$isl1
pcc <- predict.gbm(brt1, cgrid, n.trees = brt1$gbm.call$best.trees, type = "response")
# setup the storage for the predictions
pccr <- raster(cstack)
# put the values in, making sure they are numeric first
pccr <- setValues(pccr, as.numeric(pcc))
# name the new, last, layer in the stack
names(pccr) <- "brt1"

```

```{r marion to macca}
cgrid$island <- cgrid$isl2
prc <- predict.gbm(brt1, cgrid, n.trees = brt1$gbm.call$best.trees, type = "response")
# setup the storage for the predictions
prcr <- raster(cstack)
# put the values in, making sure they are numeric first
prcr <- setValues(prcr, as.numeric(prc))
# name the new, last, layer in the stack
names(prcr) <- "brt1"
```

```{r mayes to macca}
cgrid$island <- cgrid$isl3
pyc <- predict.gbm(brt1, cgrid, n.trees = brt1$gbm.call$best.trees, type = "response")
# setup the storage for the predictions
pycr <- raster(cstack)
# put the values in, making sure they are numeric first
pycr <- setValues(pycr, as.numeric(pyc))
# name the new, last, layer in the stack
names(pycr) <- "brt1"
```

##### Density predictions

Density predictions are based on presence-only data so will over-predict. We use the presence-absence predictions to mask the density prediction. Use MaxSSS as the threshold for determining present or absent in a grid cell - see Liu et al. (2013) and Guillaumot et al. (2018) for justification of using MaxSSS. Use this threshold to create a mask layer for each prediction.

```{r MaxSSS}
# code adapted from SDMPlay and dismo packages
p <- all %>% filter(pres == 1) %>% 
     predict.gbm(brt1, ., n.trees = brt1$gbm.call$best.trees, type = "response")
a <- all %>% filter(pres == 0) %>% 
     predict.gbm(brt1, ., n.trees = brt1$gbm.call$best.trees, type = "response")

eval.data <- dismo::evaluate(p = p, a = a)
tab <- base::cbind(eval.data@t, eval.data@TPR + eval.data@TNR)
maxSSS <- (base::subset(tab, tab[, 2] == max(tab[, 2])))[1, 1]
```

```{r macca to macca}
cgrid$island <- cgrid$isl1
pcc3 <- predict.gbm(brt3, cgrid, n.trees = brt3$gbm.call$best.trees, type = "response")
# setup the storage for the predictions
pccr3 <- raster(cstack)
# put the values in, making sure they are numeric first
pccr3 <- setValues(pccr, as.numeric(pcc3))
# name the new, last, layer in the stack
names(pccr3) <- "brt3"
# convert logged predictions back into unlogged
pccr3 <- exp(pccr3)

# create a presence-only mask using MaxSSS
pccr[pccr < maxSSS] <- NA
# apply mask
pccr3 <- mask(pccr3, pccr)
plot(pccr3)

# work out total estimate
ccest <- cellStats(pccr3, sum)*400
```

```{r marion to macca}
cgrid$island <- cgrid$isl2
prc3 <- predict.gbm(brt3, cgrid, n.trees = brt3$gbm.call$best.trees, type = "response")
# setup the storage for the predictions
prcr3 <- raster(cstack)
# put the values in, making sure they are numeric first
prcr3 <- setValues(prcr, as.numeric(prc3))
# name the new, last, layer in the stack
names(prcr3) <- "brt3"
# convert logged predictions back into unlogged
prcr3 <- exp(prcr3)

# create a presence-only mask using MaxSSS
prcr[prcr < maxSSS] <- NA
# apply mask
prcr3 <- mask(prcr3, prcr)
plot(prcr3)

# work out total estimate
rcest <- cellStats(prcr3, sum)*400
```

```{r mayes to macca}
cgrid$island <- cgrid$isl3
pyc3 <- predict.gbm(brt3, cgrid, n.trees = brt1$gbm.call$best.trees, type = "response")
# setup the storage for the predictions
pycr3 <- raster(cstack)
# put the values in, making sure they are numeric first
pycr3 <- setValues(pycr, as.numeric(pyc3))
# name the new, last, layer in the stack
names(pycr3) <- "brt3"
# convert logged predictions back into unlogged
pycr3 <- exp(pycr3)

# create a presence-only mask using MaxSSS
pycr[pycr < maxSSS] <- NA
# apply mask
pycr3 <- mask(pycr3, pycr)
plot(pycr3)

# work out total estimate
ycest <- cellStats(pycr3, sum)*400
```

### Figure 3

Figure 3 compares the current distributions of Blue Petrels on Macquarie Island (taken from Bird et al. 2021a) in the first panel, with BRT modelled predictions made above in the remaining panels, one for each island.

```{r}
# distribution map of surveyed densities
ggplot() +
               geom_sf(fill = 'grey', data = ccoast) +
               geom_sf(fill = NA, colour = NA, data = ccoast_buff) +
               geom_sf(data = gdat_bp, aes(fill = dens), colour = "transparent", show.legend = T) + #change to aob_d for pop dens
               scale_fill_scico(palette = "bamako", trans = "log", direction = -1) +
               scale_x_longitude(158.8, 158.9, step = 0.1) +
               theme_bw()
```

```{r}
# mac prediction
mac <- as.data.frame(pccr3, xy = T) %>% 
         drop_na()

# truncate density estimates for mapping - only pixels with >1 breeding pair predicted
macf <- mac %>% filter(layer > 0.0025) 

ggplot() +
    # set-up plot background etc.
    geom_sf(fill = 'grey', data = ccoast) +
    geom_sf(fill = NA, colour = NA, data = ccoast_buff) +
    scale_x_longitude(158.8, 158.9, step = 0.1) +
    # add GP layer
    geom_raster(data = mac, aes(x=x, y=y, fill=layer), show.legend = T) +
    scale_fill_scico(palette = "bamako", trans = "log", direction = -1) +
    theme_bw()
```

```{r}
# mar prediction
mar <- as.data.frame(prcr3, xy = T) %>% 
         drop_na()

# truncate density estimates for mapping - only pixels with >1 breeding pair predicted
marf <- mar %>% filter(layer > 0.0025) 

ggplot() +
    # set-up plot background etc.
    geom_sf(fill = 'grey', data = ccoast) +
    geom_sf(fill = NA, colour = NA, data = ccoast_buff) +
    scale_x_longitude(158.8, 158.9, step = 0.1) +
    # add GP layer
    geom_raster(data = mar, aes(x=x, y=y, fill=layer), show.legend = T) +
    scale_fill_scico(palette = "bamako", trans = "log", direction = -1) +
    theme_bw()
```

```{r}
# may prediction
may <- as.data.frame(pycr3, xy = T) %>% 
         drop_na()

# truncate density estimates for mapping - only pixels with >1 breeding pair predicted
mayf <- may %>% filter(layer > 0.0025) 

ggplot() +
    # set-up plot background etc.
    geom_sf(fill = 'grey', data = ccoast) +
    geom_sf(fill = NA, colour = NA, data = ccoast_buff) +
    scale_x_longitude(158.8, 158.9, step = 0.1) +
    # add GP layer
    geom_raster(data = may, aes(x=x, y=y, fill=layer), show.legend = T) +
    scale_fill_scico(palette = "bamako", trans = "log", direction = -1) +
    theme_bw()
```